Combining living and fossil taxa into phylogenies: the missing data issue

### 2015-01-09 Manuscript/TEM_draft.tex
Fixed comments and figures

%Packages
\usepackage{fixltx2e}
\usepackage{textcomp}
\usepackage{fullpage}
\usepackage{natbib}
\usepackage{float}
\usepackage{latexsym}
\usepackage{hyperref}
\usepackage{url}
\usepackage{hyperref}
\usepackage{epsfig}
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{bm}
\usepackage{array}
\usepackage[version=3]{mhchem}
\usepackage{ifthen}
\usepackage{caption}
\usepackage{hyperref}
\usepackage{amsthm}
\usepackage{amstext}
\usepackage{enumerate}
\usepackage[osf]{mathpazo}
\pagenumbering{arabic}

@@

\noindent{\Large \bf Effect of missing data on topological inference using a total evidence approach}
% TG: Or "Effect of missing data and inference method on topology when using a total evidence approach"
% TG: Or any other suggestions?

@@

%Authors
\noindent {\normalsize \sc Thomas Guillerme$^1$$^,$$^2$, and Natalie Cooper$^1$$^,$$^2$}\\
\noindent {\small \it 
$^1$School of Natural Sciences, Trinity College Dublin, Dublin 2, Ireland;\\
$^2$Trinity Centre for Biodiversity Research, Trinity College Dublin, Dublin 2, Ireland;}\\
\end{center}
\medskip
\noindent{\bf Corresponding author:} Thomas Guillerme, School of Natural Sciences, Trinity College Dublin, Dublin 2, Ireland; E-mail: guillert@tcd.ie; Fax: +353 1 6778094; Tel: +353 1 896 2571.\\
\vspace{1in}

@@

Although most species that have ever lived are now extinct \citep{novacek1992ext,raup1993extinction}, the majority of macroevolutionary studies focus solely on living species \citep[e.g.][]{meredithimpacts2011,jetzthe2012}. Ignoring fossil taxa may lead to misinterpretation of macroevolutionary patterns and processes such as the timing of diversification events \citep[e.g.][]{pyrondivergence2011}, relationships among lineages \citep[e.g.][]{manosphylogeny2007} or niche occupancy \citep[e.g.][]{pearmanniche2008}. This has led to increasing consensus among evolutionary biologists that fossil taxa should be included in macroevolutionary studies \citep{jacksonwhat2006,quentaldiversity2010,dietlconservation2011,slaterunifying2013,fritzdiversity2013}. However, to do this we need to be able to place living and fossil taxa into the same phylogenies; a task that remains difficult despite recent methodological developments \citep[e.g.][]{pyrondivergence2011,ronquista2012,BEASTmaster}. %Or is it cheap to cite unpublished work (Nick's proto package) in the intro?
% NC: Check Syst Biols policies, but I think it's fine as it's an available package so it'd be like citing something from GitHub. You could ask Nick before publication if there's a different citation he'd prefer you to use.

@@

In this study, we focus on the effect of missing data on our ability to recover a "best" tree topology because it is a crucial aspect of a phylogeny in many macroevolutionary studies, for example when trying to elucidate the evolutionary relationships among species \citep[e.g.][]{meredithimpacts2011,jetzthe2012}, or for studying evolutionary transitions \citep[e.g.][]{}. %CITE-check Lloyd and Friedman or birdy stuff
Although branch length estimation is also important (namely for timing extinction and/or speciation events - e.g. \citealt{ronquista2012}), we do not consider branch lengths in this study. This is partially due to difficulties with simulating branch lengths and topology simultaneously,
but also because previous studies have already empirically assessed the effect of the Total Evidence method on branch length variation but with a fixed topology approach \citep{ronquista2012,schragocombining2013,slaterphylogenetic2013,beckancient2014}.
Thus understanding the sensitivity of topology to missing data is important for assessing the accuracy of tree estimation in the Total Evidence framework. To our knowledge, this question has never been formally assessed.

@@

\begin{figure}
\centering
    \includegraphics[keepaspectratio=true]{Figures/In_main/Simulations_outline-BW.pdf}
\caption{Protocol outline.
(1) We randomly generated a birth-death tree (the "true" tree) and used it to infer a matrix with no missing data (the "complete" matrix).
(2) We removed data from the morphological part of the "complete" matrix resulting in 125 "missing-data" matrices.
(3) We built phylogenetic trees from each matrix using both Maximum Likelihood and Bayesian methods.
(4) We compared the "missing-data" trees to the "best" tree.
We repeated these four steps 50 times.}
\label{Fig_Outline}
\end{figure}
% NC: Can you centre these around the boxes? and make the comparing topologies arrow touch the two triangles? I'll show you what I mean. 
% NC: Also check the figure is in a format and of correct quality for Syst Biol.

@@

\subsection{1-Generating the matrix}
First we randomly generated a "true" tree of 50 taxa in R v. 3.0.2 \citep{R302} using the package diversitree v. 0.9-6 \citep{fitzjohndiversitree2012}. We generated the tree using a birth death process by sampling speciation ($\lambda$) and extinction ($\mu$) rates from a uniform distribution but maintaining $\lambda$ $>$ $\mu$ \citep{paradistime-dependent2011}. We implemented a rejection sampling algorithm to select only trees with 25 living and 25 fossil taxa to ensure that we had enough taxa of each type for our missing data simulations to work. We then added an outgroup to the tree, using the mean branch length of the tree to separate the outgroup from the rest of the taxa, and with the branch length leading to the outgroup set as the sum of the mean branch length and the longest root-to-tip length of the tree.

@@

\subsubsection{Maximum Likelihood}
The "best" tree and the "missing-data" trees were inferred using RAxML v. 8.0.20 \citep{Stamatakis21012014}. For the molecular data, we used the GTR + $\Gamma_4$ model (\citealt{tavare1986}; default GTRGAMMA in RAxML v. 8.0.20; \citealt{Stamatakis21012014}).% as a generalization of the HKY + $\Gamma_4$ model \citep{HKY85}. The GTR model can be seen as a generalization of the HKY model (the two parameters from the HKY model are implicitly included in the six from GTR model - \citealt{stamatakisa2008}).
For the morphological data, we used the implemented Markov \textit{k} state model \citep{lewisa2001} %which is a generalization of the JC69 model \citep{jc69} with \textit{k} $\geq$ 2
 assuming an equal state frequency and a unique overall substitution rate ($\mu$) following a gamma distribution of the rate variation with four distinct categories (M\textit{k} + $\Gamma_4$; -K MK option in RAxML v. 8.0.20; \citealt{Stamatakis21012014}).

To measure the phylogenetic signal of our simulations, we first ran a fast bootstrap analysis (Lazy Sub-tree Rearrangement) with 500 replicates on the "complete" matrix. We removed all the simulations with a median bootstrap support lower than 50 as a proxy for weak phylogenetic signal \citep{zanderminimal2004}. We repeated this selection until we obtained 50 sets of simulations (i.e. 50 "complete" and 50 x 125 "missing-data" matrices) with a relative good phylogenetic signal (median bootstrap $>$ 50). On these selected simulations, we used the fast bootstrap algorithm and performed 1000 bootstraps for each tree inference to assess topological support \citep{pattengale2010many}.
%The bootstrap algorithm used in RAxML is the Lazy Sub-tree Rearrangement (LSR) which consists of pruning one sub-tree from the tree and subsequently reinserting it to all neighbouring branches \citep{stamatakisa2008}. Sub-tree Pruning and Reinserting methods (SPR) have been demonstrated to be better than other methods (e.g. Nearest Neighbouring Interchange - NNI) for recovering good bootstrap values \citep{salamin2003}.
Using these parameters took \texttildelow 8 CPU years to build 50 sets of 125 bootstrapped Maximum Likelihood trees (2.30GHz clock speed nodes).

@@

\subsubsection{Robinson-Foulds distance}
The Robinson-Foulds distance \citep{RF1981}, or "path difference", measures the number of shared clades across two trees. The metric reflects the distance between the distributions of tips among clades in the two trees (\citealt{RF1981}; see \hyperref[SupplementaryMaterial]{Supplementary Material S2} for calculation details). This metric is bounded between one, when the two trees are identical, and $n-2$ (for two trees with $n$ taxa) when there is no shared clade in the two trees. This metric is sensitive to minor changes in clade conservation: if the trees are composed of two clades of three taxa (\textit{(((a,b),c),((d,e),f))}), the swapping of any two taxa will lead to a maximal score of the Robinson-Foulds distance indicating poor tree similarity. We normalised this metric by % NC: DETAILS 
\citep{Bogdanowicz2012}. This allows us to compare the metric for any number of taxa by scaling it with the expected value of the metric when comparing two random trees (see \hyperref[SupplementaryMaterial]{Supplementary Material S2} for the calculation details). Therefore, when the Normalised Robinson-Foulds distance between two trees is equal to one, the trees are identical; if the distance is equal to zero, the trees are no more different than expected by chance; finally if the distance is less than zero, the trees are more different than expected by chance.

@@

\subsubsection{Paired trees comparisons}
\label{tree_comparisons}
For the Maximum Likelihood and Bayesian consensus trees we performed pairwise comparisons between the "best" tree and each "missing-data" tree using both the Normalised Robinson-Foulds and Normalised Triplets metrics with the TreeCmp java script \citep{Bogdanowicz2012} resulting each time in 125 Normalised Robison-Foulds distances and 125 Normalised Triplets distance. %For each metric, we then normalized the value using the Normalized Tree Similarity scaled by the mean value of 1000 pairwise random tree comparisons for the metric in question and n = 51 taxa (see \hyperref[SupplementaryMaterial]{Supplementary Material Section 2}).
%We compared each "missing-data" tree with the "best" tree for each of our 50 simulation runs resulting in 50 comparisons for each "missing-data" tree. We calculated the mode and the 50\% and 95\% confidence intervals from the resulting distribution using the hdrcde R package v3.1 \citep{hdrcde}.
Also, to take into account the uncertainty of tree inference in both methods cited above, instead of using the single Maximum Likelihood and Bayesian consensus trees, we extracted 1000 random bootstrapped trees from the Maximum Likelihood analysis and 1000 from the posterior tree distribution of the Bayesian analysis for the "best" trees and then did the same for the 125 "missing data" trees (resulting in 1000 "best" trees and 125$\times$1000 "missing data" trees). 
For a given set of 1000 "missing data" trees and the 1000 "best" trees, we sampled one "missing data" tree and one "best" tree at random and compared them using both the Normalised Robinson-Foulds and Normalised Triplets metrics as described above. We repeated this 1000 times for each set of "missing data" trees resulting in 125$\times$1000 values for each metric. We repeated all the paired trees comparisons described above for each of the 50 simulations. We then calculated the mode and the 50\% and 95\% confidence intervals from the resulting distribution using the hdrcde R package v. 3.1 \citep{hdrcde}.

\subsection{Testing the effects of the missing data parameters on topological recovery} % NC: I think this is really a separate section to tree comparison.
Finally, we tested the effects of our missing data parameters ($M_{L}$, $M_{F}$, $M_{C}$ and their interactions) on our ability to recover the "best" tree topology in a Total Evidence framework. We also assessed the effect of our missing data parameters jointly with the effects of different tree inference and uncertainty methods (i.e. Maximum Likelihood, Bayesian consensus, Maximum Likelihood bootstrap trees and Bayesian posterior tree distribution).

We measured similarities among the distributions of the different metrics scores (Normalised Robinson-Foulds and Normalised Triplets distance) using the Bhattacharyya Coefficient \citep{Bhattacharyya}. The Bhattacharyya Coefficient is the probability of overlap between two distributions (\citealt{Bhattacharyya}; see \hyperref[SupplementaryMaterial]{Supplementary Material Section 2} for calculation details). Note that this is  equivalent to performing a two-sided t-test, but we use the Bhattacharyya Coefficient here because we are comparing distributions not means. To assess the effect of our missing data parameters, we calculated the Bhattacharyya Coefficient between the distributions of the different metrics scores (Normalised Robinson-Foulds and Normalised Triplets distance) for each pairwise combination of missing data parameters ($M_{L}$, $M_{F}$, $M_{C}$) and parameter states (0\%, 10\%, 25\%, 50\%, 75\%), i.e. $M_{L}$ = 0\%, $M_{F}$ = 0\%, $M_{C}$ = 0\%; $M_{L}$ = 10\%, $M_{F}$ = 0\%, $M_{C}$ = 0\% etc. (see Figure % FIGURE
for more details). This resulted in 7875 pairwise comparisons (a triangular matrix with $3^5$$\times$$3^5$ cells). We performed this procedure separately for each tree inference and uncertainty method. When two combinations of missing data parameters have a similar ability to recover the "best" tree topology the Bhattacharyya Coefficient will be close to one. Conversely, if the two combinations of missing data parameters differ, the Bhattacharyya Coefficient will be close to zero.

% NC: Maybe put this in the results?
% We represented the results in a triangular matrix with the values of each pairwise Bhattacharyya Coefficient coloured according to their values (in green when the distribution overlap completely (Bhattacharyya Coefficient = 1) and in red when they don't (Bhattacharyya Coefficient = 0) - Fig.~\ref{Fig_Results-paircomp_within}).

To assess the effect of the different tree inference and uncertainty methods (i.e. Maximum Likelihood, Bayesian consensus, Maximum Likelihood bootstrap trees and Bayesian posterior tree distribution) on our ability to recover the "best" tree topology, we calculated the Bhattacharyya Coefficient between the distributions of the different metrics scores (Normalised Robinson-Foulds and Normalised Triplets distance) for each pairwise combination of tree inference and uncertainty methods i.e. Maximum Likelihood \textit{versus} Bayesian consensus; Maximum Likelihood \textit{versus} Maximum Likelihood bootstrap trees etc. (see Figure % FIGURE
for more details). Note that this procedure pools results from across all missing data parameter combinations so it results in just six pairwise comparisons. When two tree inference or uncertainty methods have a similar ability to recover the "best" tree topology the Bhattacharyya Coefficient will be close to one. Conversely, if the two tree inference or uncertainty methods differ, the Bhattacharyya Coefficient will be close to zero.

@@

\section{Results}

% TG: Ok I removed all the figures + table on top of the section in here and all the text down. We might to have a think of where to anchor the figures later on.

\begin{figure} 
\centering
    \includegraphics[width=1\textwidth]{Figures/In_main/AllMethods-RF+Tr-colour.pdf}
\caption{The effects of increasing missing data on topological recovery using Maximum Likelihood trees (black), Bayesian consensus trees (red), Maximum Likelihood bootstrap trees (green) and Bayesian posterior tree distributions (blue). The percentage of missing data for each parameter ($M_{L}$, $M_{F}$ and $M_{C}$) is shown on the x axis. Topological recovery was measured using two different tree comparison metrics: Normalised Robinson-Foulds distance (upper row) and Normalised Triplets distance (lower row). The graph shows the modal value (points), the 50\% (thick solid lines) and 95\% (thin dashed lines) confidence intervals of the distributions of the tree comparison metric for each missing data parameter and tree inference method. }
\label{Fig_Results-permeth_perparam} %Differences within a subset of parameters and between methods
\end{figure}

% NC: Y axes should be standardised Robinson Foulds
% NC: X-axes should have (Mc) etc in brackets after the names
% NC: Also should probably have the X axes read : Missing character (Mc, %) and remove % from tick labels
% TG: Ok, will fix later on since it's all scripted (no illustrator cheats)

%NC: I would rewrite these figure labels as follows. I prefer the details of colours and things to be in brackets not in the text. Rewrite the others in a similar fashion.
%NC: Remeber it's not topology, but topological recovery that you are measuring
%The effects of increasing missing data on topological recovery using Maximum Likelihood trees (black), Bayesian consensus trees (red), Maximum Likelihood bootstrap trees (green) and Bayesian posterior tree distributions (blue). The percentage of missing data for each parameter ($M_{L}$, $M_{F}$ and $M_{C}$) is shown on the x axis. Topological recovery was measured using two different tree comparison metrics: Normalised Robinson-Foulds distance (upper row) and Normalised Triplets distance (lower row). The graph shows the modal value (points), the 50\% (thick solid lines) and 95\% (thin dashed lines) confidence intervals of the distributions of the tree comparison metric for each missing data parameter and tree inference method.

@@

%Hide this figure for 'fast' build
%\begin{figure} 
%\centering
%    \includegraphics[width=1\textwidth]{Figures/In_main/PairwiseComp-Baycon-RF+Tr-colour.pdf}
%\caption{Effect of missing data on topological recovery using Bayesian consensus trees. The percentage of missing data (gradient colour from 0\% (white), 10\%, 25\%, 50\% and 75\% (black)) for each parameter ($M_{L}$ (upper line), $M_{F}$ (middle line) and $M_{C}$(lower line)) is shown on both x and y axis. Topological recovery is represented by the probability of the two different tree comparison metrics (Normalised Robinson-Foulds distance (A - right) and Normalised Triplets distance (B - left)) distribution overlap with the "best" tree using the the Bhattacharyya Coefficients (low probability of overlap - red ; high probability of overlap - green).}
\label{Fig_Results-paircomp_within}
%\end{figure} %Pairwise BC for the Bayesian consensus trees for RF and TR. (parameters differences)

>>>>

%Packages
\usepackage{pdflscape}
\usepackage{fixltx2e}
\usepackage{textcomp}
\usepackage{fullpage}
\usepackage{natbib}
\usepackage{float}
\usepackage{latexsym}
\usepackage{hyperref}
\usepackage{url}
\usepackage{hyperref}
\usepackage{epsfig}
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{bm}
\usepackage{array}
\usepackage[version=3]{mhchem}
\usepackage{ifthen}
\usepackage{caption}
\usepackage{hyperref}
\usepackage{amsthm}
\usepackage{amstext}
\usepackage{enumerate}
\usepackage[osf]{mathpazo}
\pagenumbering{arabic}

@@

\noindent{\Large \bf Effect of missing data on topological inference using a total evidence approach}

@@

\noindent {\normalsize \sc Thomas Guillerme$^1$$^,$$^2$, and Natalie Cooper$^1$$^,$$^2$}\\
\noindent {\small \it 
$^1$School of Natural Sciences, Trinity College Dublin, Dublin 2, Ireland;\\
$^2$Trinity Centre for Biodiversity Research, Trinity College Dublin, Dublin 2, Ireland;}\\
\end{center}
\medskip
\noindent{\bf Corresponding author:} Thomas Guillerme, School of Natural Sciences, Trinity College Dublin, Dublin 2, Ireland; E-mail: guillert@tcd.ie; Fax: +353 1 6778094; Tel: +353 1 896 2571.\\
\vspace{1in}

@@

Although most species that have ever lived are now extinct \citep{novacek1992ext,raup1993extinction}, the majority of macroevolutionary studies focus solely on living species \citep[e.g.][]{meredithimpacts2011,jetzthe2012}. Ignoring fossil taxa may lead to misinterpretation of macroevolutionary patterns and processes such as the timing of diversification events \citep[e.g.][]{pyrondivergence2011}, relationships among lineages \citep[e.g.][]{manosphylogeny2007} or niche occupancy \citep[e.g.][]{pearmanniche2008}. This has led to increasing consensus among evolutionary biologists that fossil taxa should be included in macroevolutionary studies \citep{jacksonwhat2006,quentaldiversity2010,dietlconservation2011,slaterunifying2013,fritzdiversity2013}. However, to do this we need to be able to place living and fossil taxa into the same phylogenies; a task that remains difficult despite recent methodological developments \citep[e.g.][]{pyrondivergence2011,ronquista2012,BEASTmaster}.

@@

In this study, we focus on the effect of missing data on our ability to recover a "best" tree topology because it is a crucial aspect of a phylogeny in many macroevolutionary studies, for example when trying to elucidate the evolutionary relationships among species \citep[e.g.][]{meredithimpacts2011,jetzthe2012}, or for studying evolutionary transitions \citep[e.g.][]{friedmanexplosive2010}. Although branch length estimation is also important (namely for timing extinction and/or speciation events - e.g. \citealt{ronquista2012}), we do not consider branch lengths in this study. This is partially due to difficulties with simulating branch lengths and topology simultaneously, but also because previous studies have already empirically assessed the effect of the Total Evidence method on branch length variation but with a fixed topology approach \citep{ronquista2012,schragocombining2013,slaterphylogenetic2013,beckancient2014}. Thus understanding the sensitivity of topology to missing data is important for assessing the accuracy of tree estimation in the Total Evidence framework. To our knowledge, this question has never been formally assessed.

@@

\begin{figure}
\centering
    \includegraphics[keepaspectratio=true]{Figures/In_main/Simulations_outline-BW.pdf}
\caption{Protocol outline.
(1) We randomly generated a birth-death tree (the "true" tree) and used it to infer a matrix with no missing data (the "complete" matrix).
(2) We removed data from the morphological part of the "complete" matrix resulting in 125 "missing-data" matrices.
(3) We built phylogenetic trees from each matrix using both Maximum Likelihood and Bayesian methods.
(4) We compared the "missing-data" trees to the "best" tree.
We repeated these four steps 50 times.}
\label{Fig_Outline}
\end{figure}

@@

\subsection{1-Generating the matrix}
\label{Generating_the_matrix}
First we randomly generated a "true" tree of 50 taxa in R v. 3.0.2 \citep{R302} using the package diversitree v. 0.9-6 \citep{fitzjohndiversitree2012}. We generated the tree using a birth death process by sampling speciation ($\lambda$) and extinction ($\mu$) rates from a uniform distribution but maintaining $\lambda$ $>$ $\mu$ \citep{paradistime-dependent2011}. We implemented a rejection sampling algorithm to select only trees with 25 living and 25 fossil taxa to ensure that we had enough taxa of each type for our missing data simulations to work. We then added an outgroup to the tree, using the mean branch length of the tree to separate the outgroup from the rest of the taxa, and with the branch length leading to the outgroup set as the sum of the mean branch length and the longest root-to-tip length of the tree.

@@

\subsubsection{Maximum Likelihood}
The "best" tree and the "missing-data" trees were inferred using RAxML v. 8.0.20 \citep{Stamatakis21012014}. For the molecular data, we used the GTR + $\Gamma_4$ model (\citealt{tavare1986}; default GTRGAMMA in RAxML v. 8.0.20; \citealt{Stamatakis21012014}). For the morphological data, we used the implemented Markov \textit{k} state model \citep{lewisa2001} assuming an equal state frequency and a unique overall substitution rate ($\mu$) following a gamma distribution of the rate variation with four distinct categories (M\textit{k} + $\Gamma_4$; -K MK option in RAxML v. 8.0.20; \citealt{Stamatakis21012014}).

To measure the phylogenetic signal of our simulations, we first ran a fast bootstrap analysis (Lazy Sub-tree Rearrangement) with 500 replicates on the "complete" matrix. We removed all the simulations with a median bootstrap support lower than 50 as a proxy for weak phylogenetic signal \citep{zanderminimal2004}. We repeated this selection until we obtained 50 sets of simulations (i.e. 50 "complete" and 50 x 125 "missing-data" matrices) with a relative good phylogenetic signal (median bootstrap $>$ 50). On these selected simulations, we used the fast bootstrap algorithm and performed 1000 bootstraps for each tree inference to assess topological support \citep{pattengale2010many}. Using these parameters took \texttildelow 8 CPU years to build 50 sets of 125 bootstrapped Maximum Likelihood trees (2.30GHz clock speed nodes).

@@

\subsubsection{Robinson-Foulds distance}
The Robinson-Foulds distance \citep{RF1981}, or "path difference", measures the number of shared clades across two trees. The metric reflects the distance between the distributions of tips among clades in the two trees (\citealt{RF1981}; see \hyperref[SupplementaryMaterial]{Supplementary Material S2} for calculation details). This metric is bounded between one, when the two trees are identical, and $n-2$ (for two trees with $n$ taxa) when there is no shared clade in the two trees. This metric is sensitive to minor changes in clade conservation: if the trees are composed of two clades of three taxa (\textit{(((a,b),c),((d,e),f))}), the swapping of any two taxa will lead to a maximal score of the Robinson-Foulds distance indicating poor tree similarity. We normalised this metric by divided the Robinson-Foulds distance by the mean Robinson-Foulds distance of 1000 random trees comparisons \citep{Bogdanowicz2012}. This allows us to compare the metric for any number of taxa by scaling it with the expected value of the metric when comparing two random trees (see \hyperref[SupplementaryMaterial]{Supplementary Material S2} for the calculation details). Therefore, when the Normalised Robinson-Foulds distance between two trees is equal to one, the trees are identical; if the distance is equal to zero, the trees are no more different than expected by chance; finally if the distance is less than zero, the trees are more different than expected by chance.

@@

\subsubsection{Paired trees comparisons}
\label{tree_comparisons}
For the Maximum Likelihood and Bayesian consensus trees we performed pairwise comparisons between the "best" tree and each "missing-data" tree using both the Normalised Robinson-Foulds and Normalised Triplets metrics with the TreeCmp java script \citep{Bogdanowicz2012} resulting each time in 125 Normalised Robison-Foulds distances and 125 Normalised Triplets distance. Also, to take into account the uncertainty of tree inference in both methods cited above, instead of using the single Maximum Likelihood and Bayesian consensus trees, we extracted 1000 random bootstrapped trees from the Maximum Likelihood analysis and 1000 from the posterior tree distribution of the Bayesian analysis for the "best" trees and then did the same for the 125 "missing data" trees (resulting in 1000 "best" trees and 125$\times$1000 "missing data" trees). 
For a given set of 1000 "missing data" trees and the 1000 "best" trees, we sampled one "missing data" tree and one "best" tree at random and compared them using both the Normalised Robinson-Foulds and Normalised Triplets metrics as described above. We repeated this 1000 times for each set of "missing data" trees resulting in 125$\times$1000 values for each metric. We repeated all the paired trees comparisons described above for each of the 50 simulations. We then calculated the mode and the 50\% and 95\% confidence intervals from the resulting distribution using the hdrcde R package v. 3.1 \citep{hdrcde}.

\subsection{Testing the effects of the missing data parameters on topological recovery}
Finally, we tested the effects of our missing data parameters ($M_{L}$, $M_{F}$, $M_{C}$ and their interactions) on our ability to recover the "best" tree topology in a Total Evidence framework. We also assessed the effect of our missing data parameters jointly with the effects of different tree inference and uncertainty methods (i.e. Maximum Likelihood, Bayesian consensus, Maximum Likelihood bootstrap trees and Bayesian posterior tree distribution).

We measured similarities among the distributions of the different metrics scores (Normalised Robinson-Foulds and Normalised Triplets distance) using the Bhattacharyya Coefficient \citep{Bhattacharyya}. The Bhattacharyya Coefficient is the probability of overlap between two distributions (\citealt{Bhattacharyya}; see \hyperref[SupplementaryMaterial]{Supplementary Material Section 2} for calculation details). Note that this is  equivalent to performing a two-sided t-test, but we use the Bhattacharyya Coefficient here because we are comparing distributions not means. To assess the effect of our missing data parameters, we calculated the Bhattacharyya Coefficient between the distributions of the different metrics scores (Normalised Robinson-Foulds and Normalised Triplets distance) for each pairwise combination of missing data parameters ($M_{L}$, $M_{F}$, $M_{C}$) and parameter states (0\%, 10\%, 25\%, 50\%, 75\%), i.e. $M_{L}$ = 0\%, $M_{F}$ = 0\%, $M_{C}$ = 0\%; $M_{L}$ = 10\%, $M_{F}$ = 0\%, $M_{C}$ = 0\% etc. (see Fig.~\ref{Fig_Bhattacharyya_Coefficients} for more details). This resulted in 7875 pairwise comparisons (a triangular matrix with $3^5$$\times$$3^5$ cells). We performed this procedure separately for each tree inference and uncertainty method. When two combinations of missing data parameters have a similar ability to recover the "best" tree topology the Bhattacharyya Coefficient will be close to one. Conversely, if the two combinations of missing data parameters differ, the Bhattacharyya Coefficient will be close to zero.

To assess the effect of the different tree inference and uncertainty methods (i.e. Maximum Likelihood, Bayesian consensus, Maximum Likelihood bootstrap trees and Bayesian posterior tree distribution) on our ability to recover the "best" tree topology, we calculated the Bhattacharyya Coefficient between the distributions of the different metrics scores (Normalised Robinson-Foulds and Normalised Triplets distance) for each pairwise combination of tree inference and uncertainty methods i.e. Maximum Likelihood \textit{versus} Bayesian consensus; Maximum Likelihood \textit{versus} Maximum Likelihood bootstrap trees etc. (see Fig.~\ref{Fig_Bhattacharyya_Coefficients} for more details). Note that this procedure pools results from across all missing data parameter combinations so it results in just six pairwise comparisons. When two tree inference or uncertainty methods have a similar ability to recover the "best" tree topology the Bhattacharyya Coefficient will be close to one. Conversely, if the two tree inference or uncertainty methods differ, the Bhattacharyya Coefficient will be close to zero.

\begin{figure} 
\centering
    \includegraphics[width=1\textwidth]{Figures/In_main/Bhattacharyya_Coefficients.pdf}
\caption{Bhattacharyya Coefficient calculation outline. A, B and C are distribution of tree similarity (Normalised Robinson-Foulds or Normalised Triplets distances) for any combination of missing data parameters (e.g. $M_{L}$ = 10\%, $M_{F}$ = 50\%, $M_{C}$ = 75\%). \textbf{(x)} and \textbf{(y)} are two different tree inference methods (e.g. Maximum Likelihood or Bayesian). 1: the Bhattacharyya Coefficient (BC) is the overlap of the distribution of tree similarity between two combination of missing data parameters. 2: the Bhattacharyya Coefficient (BC) is the overlap of the distribution of tree similarity between two methods for the same combination of missing data parameters.}
\label{Fig_Bhattacharyya_Coefficients} %Bhattacharyya coefficients calculation outline
\end{figure}

@@

\section{Results}
As the amount of missing data in the morphological part of the Total Evidence matrix increases, our ability to recover the "best" tree topology decreases, regardless of the missing data parameter ($M_{L}$, $M_{F}$ or $M_{C}$), the tree inference method (Maximum Likelihood or Bayesian) or the tree comparison metric used (Normalised Robinson-Foulds or Normalised Triplets distance). However, the different missing data parameters and tree inference methods do not affect the topology in the same way (Fig.~\ref{Fig_Results-permeth_perparam} and Fig.~\ref{Fig_Results-global_perparam}).

\begin{figure} 
\centering
    \includegraphics[width=1\textwidth]{Figures/In_main/AllMethods-RF+Tr-colour.pdf}
\caption{The effects of increasing missing data on topological recovery using Maximum Likelihood trees (black), Bayesian consensus trees (blue), Maximum Likelihood bootstrap trees (orange) and Bayesian posterior tree distributions (blue). The percentage of missing data for each parameter ($M_{L}$, $M_{F}$ and $M_{C}$) is shown on the x axis. Topological recovery was measured using two different tree comparison metrics: Normalised Robinson-Foulds distance (upper row) and Normalised Triplets distance (lower row). The graph shows the modal value (points), the 50\% (thick solid lines) and 95\% (thin dashed lines) confidence intervals of the distributions of the tree comparison metric for each missing data parameter and tree inference method.}
\label{Fig_Results-permeth_perparam} %Differences within a subset of parameters and between methods
\end{figure}

@@

\subsection{Individual effects of missing data parameters}
As the amount of missing data increases across all three parameters, our ability to recover the "best" tree topology decreases (Fig.~\ref{Fig_Results-permeth_perparam}).
The Normalised Robinson-Foulds distance is always lower for the Maximum Likelihood trees than for the Bayesian consensus trees (median Bhattacharrya Coefficient = 0.69, 0.48 and 0.66 for $M_{L}$, $M_{F}$ and $M_{C}$ respectively; Fig.~\ref{Fig_Results-permeth_perparam}; \hyperref[SupplementaryMaterial]{Supplementary Material Section 3} Table 5, 6 and 7). %add link?
However, The Normalised Triplets distance is similar between the Maximum Likelihood trees and the Bayesian consensus trees for all the parameters ($M_{L}$, $M_{F}$ and $M_{C}$) (median Bhattacharrya Coefficient = 0.84, 0.75 and 0.80 for $M_{L}$, $M_{F}$ and $M_{C}$ respectively; Fig.~\ref{Fig_Results-permeth_perparam}; \hyperref[SupplementaryMaterial]{Supplementary Material Section 3} Table 5, 6 and 7). %add link?

\subsection{Combined effect of missing data parameters}
As expected, our ability to recover the "best" tree topology is worst when each parameter contains the maximum amount of missing data (i.e. $M_{L}$ = 75\%, $M_{F}$ = 75\% and $M_{C}$ = 75\%), and best when there is no missing data (i.e. $M_{L}$ = 0\%, $M_{F}$ = 0\%, $M_{C}$ = 0\%; Fig.~\ref{Fig_Results-global_perparam}; \hyperref[SupplementaryMaterial]{Supplementary Material Section 3}). Figure ~\ref{Fig_Results-paircomp_within} shows the similarity of distributions of tree metrics in a triangular matrix with the values of each pairwise Bhattacharyya Coefficient coloured according to their values (orange when the distributions overlap completely, Bhattacharyya Coefficient = 1, and blue when they do not, Bhattacharyya Coefficient = 0; Fig.~\ref{Fig_Results-paircomp_within}).

\begin{figure} 
\centering
     %Hide this figure for 'fast' build
%    \includegraphics[width=1\textwidth]{Figures/In_main/PairwiseComp-Baycon-RF+Tr-colour.pdf}
    \includegraphics[width=1\textwidth]{Figures/In_main/PairwiseComp-Baycon-RF+Tr-colour.png} %bitmap version - 300 dpi RGB
\caption{Effect of missing data on topological recovery using Bayesian consensus trees. The percentage of missing data (gradient colour from 0\% (white), 10\%, 25\%, 50\% and 75\% (black)) for each parameter ($M_{L}$ (upper line), $M_{F}$ (middle line) and $M_{C}$(lower line)) is shown on both x and y axis. Topological recovery is represented by the probability of the two different tree comparison metrics (Normalised Robinson-Foulds distance (A - right) and Normalised Triplets distance (B - left)) distribution overlap with the "best" tree using the the Bhattacharyya Coefficients (low probability of overlap - blue ; high probability of overlap - orange).}
\label{Fig_Results-paircomp_within}
\end{figure} %Pairwise BC for the Bayesian consensus trees for RF and TR. (parameters differences)

Using both Normalised Robinson-Foulds and Normalised Triplets distances from the Bayesian consensus trees, the parameter combination with no missing data (i.e. $M_{L}$ = 0\%, $M_{F}$ = 0\%, $M_{C}$ = 0\%) is always the most dissimilar to all the other parameter combinations (Fig.~\ref{Fig_Results-paircomp_within}). However, the Normalised Robinson-Foulds distance (median Bhattacharrya coefficient = 0.79 ; Fig.~\ref{Fig_Results-paircomp_within}-A) displays more dissimilarities than the Normalised Triplets distance (median Bhattacharrya coefficient = 0.81; Fig.~\ref{Fig_Results-paircomp_within}-B).Additionally, when using the Normalised Robinson-Foulds distance, once $M_{L}$ $\geq$  50\%, there is no additional affect of $M_{F}$ and $M_{C}$, regardless of the amount of missing data in these parameters (Fig.~\ref{Fig_Results-paircomp_within}-A). Likewise, once $M_{C}$ $\geq$ 50\%, there is no additional affect of $M_{L}$ and $M_{F}$ (Fig.~\ref{Fig_Results-paircomp_within}-A).

For all combinations of missing data parameters and tree comparison metrics, the Maximum Likelihood bootstrap trees and the Bayesian posterior tree distributions perform very similarly (median Bhattacharrya Coefficient = 0.85 and 0.98, using Normalised Robinson-Foulds distance or Normalised Triplets distance respectively; see Table ~\ref{Tab_Results-Difference_methods}). However, these two methods perform worse than the Bayesian consensus trees sing Normalised Robinson-Foulds distance (median Bhattacharrya Coefficient = 0 and 0.01, for the Maximum Likelihood bootstrap trees and the Bayesian posterior tree distribution respectively; see Table ~\ref{Tab_Results-Difference_methods}; Fig.~\ref{Fig_Results-permeth_perparam} and \hyperref[SupplementaryMaterial]{Supplementary Material Section 3}).
